A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-08-15 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: Align dice example with book
Instead of e_123456 = e_1 + e_23456 we split into e_3 + e_12456 like we do in the book.

In the past few decades, there has been a growing interest in developing approaches that generalize the notion of states and physical theories in terms of probabilistic and/or entropic concepts, [CITE - GPT, Ariel, Reginatto/Hall...] which stems from the underlying aim to characterize physical theories from an operational viewpoint. One basic insight is that statistical/probabilistic concepts often provide a suitable basis for this generalization, and it is in this setting where most of these approaches operate. A key problem, then, is what notion of probability should be used given that the standard measure theoretic Kolmogorovian approach does not generally work in quantum mechanics.[CITE] Quasiprobabilistic approaches [CITE] are also used in quantum mechanics, but it is not completely clear how exactly they relate to standard one.

@@

The starting point is that any physical law deals with experimentally reproducible relationships. This implies a physical theory must give us the space of statistical ensembles (i.e. the collection of all possible preparation procedures) that the theory allows. As a basic requirement, ensembles must be experimentally well-defined (leading to a topological space), allow statistical mixing (leading to a convex structure) and quantify the variability of the elements of the ensembles (leading to an entropy function).\footnote{The main point of departure from GPTs is the entropy requirement. We found that a convex set is not enough to fully characterize an ensemble space, especially in the infinite dimensional case.} Additionally, though not necessary, we require having enough real valued quantities to identify each ensemble (i.e. there is a family of affine functions that identify each ensemble through their values). Mathematically, these premises lead us to assume that an ensemble space is a convex subset of a metrizable locally convex topological vector space. The question is, then, what measure theoretic tool can quantify the number of states (as in the classical Liouville measure) or the number of possible outcomes of a measurements (as in the dimension of orthogonal subspaces of an operator)? What measure theoretic tool can represent each ensemble? If we are to define a physically meaningful notion of probability density, both of these questions must be answered.

We will first review how measure theory is used in classical probability [cite] and classical statistical mechanics[cite]. We will connect this to Choquet theory[cite phelps], which aims to represent each point of a convex subset as a probability measure over its extreme points (i.e. each ensemble can be represented by a probability measure over pure states). While such representation is always possible, at least in the compact case, it is not unique. Uniqueness requires the space to be a Choquet simplex, the generalization of the finite-dimensional simplex, and represents the classical case. The key insight, then, is that an ensemble space is classical if and only if each ensembles admits a unique decomposition in terms of pure states.

We will then review some results from the use of quasi-probability measures. The most used is the Wigner function,[cite] that represents each state as a signed probability measure over phase space, and expectations of observables are recovered through normal integral. While the Wigner function represents the most known instance, there are different choices depending on the problem at hand.[cite] The key insight is that it is not negative probability that characterizes quasi-probability,\footnote{Some representation guarantee the quasi-probability to be non-negative.} but the lack of mutual exclusivity of the points of the space. The quasi-probability measures, in fact, are not defined over the extreme points, but over a suitable set $K$ such that the space of signed probability measures is isomorphic as a vector space to the one that embeds the ensemble space. Therefore, while signed measures are a powerful tool for calculation in specific problems, they cannot give us a general principled representation of the physical theory.

The above two approaches exhaust what can be done with additive measures, so we turn to the non-additive case. Given an ensemble $\ens$, we can ask what fraction can be produced by a mixture of a given ensembles. This leads to the fraction capacity: a sub-additive set function that, like the probability measure, is bounded to one, monotonic and continuous from above and below. Over the extreme points, the fraction capacity represents the supremum of all possible measures that can represent the same ensemble in Choquet theory. The classical case exactly corresponds to the simplex case, in which fraction capacity corresponds to a unique measure, and is therefore additive. This again solidifies the connection between classical probability, additivity of measures and mutual exclusivity.

We can also generalize a notion of count of states. While in classical statistical mechanics we start with a measure that quantifies the count of states and define the entropy using its logarithm, for the ensemble space we proceed in the opposite way: we start from the entropy and we define the count of states as the exponential of the entropy (i.e. the entropic volume). From this, we can define a state capacity, which is again a sub-additive measure. It reduces to the Liouville measure in the classical case and the dimensionality of subspaces in the quantum case.

Lastly, we lay down the requirements for a non-additive generalization of probability theory, which should be in principle possible. Since there is a wide literature on the subject, [cite] we only show that the most used non-additive integrals, the Sugeno and Choquet integral, do not work for our purposes. This means that more work will be needed to develop the appropriate tools.

A final remark before we start. Too often works in the foundations of physics focus on a particular aspect without concern to a broader connection to all relevant mathematical, physical and philosophical aspects, such as information geometry, functional analysis, quantum logic and so on. However, such a complete work would be effectively overwhelming both to produce and to read. The main part of the paper, then, highlights the main conceptual results regarding measure theoretic representations, the appendix goes through the relevant mathematical detail while all the rest is left to our evolving collaborative open source/access work.[cite our book]

@@

\textbf{Fraction capacity.} Let us concentrate on the following problem: given an ensemble $\ens \in \Ens$ and a (Borel) set of ensembles $A \subseteq \Ens$, what fraction of $\ens$ can be constructed by a mixture of $A$? How much of $\ens$ can be explained as coming from preparations corresponding to $A$? For example, let $\Ens$ be the space of all probability distributions for a six-sided die. Let $\ens_{123456}$ be the uniform distribution over all outcomes. Let $A_1 = \{\ens_1\}$ where $\ens_1$ represents outcome one with 100\% probability. Since we can write $\ens_{123456} = \frac{1}{6} \ens_{1} + \frac{5}{6} \ens_{23456}$, where $\ens_{23456}$ represents a uniform distributions over the five outcomes, $1/6$ of $\ens$ can be constructed from $A$, but no more. Similarly, if $A_{12} = \{\ens_{1},\ens_{2}\}$, we can write $\ens = \frac{1}{3} \left(\frac{1}{2} \ens_1 + \frac{1}{2} \ens_2 \right)  + \frac{2}{3} \ens_{3456}$, so $1/3$ of $\ens$ can be constructed from $A$, but no more. Note that, given the uniform distribution, $1/6$ is exactly the probability for the event $A_1$ and $1/3$ the probability for event $A_{12}$. This gives the basic insight for our definitions.

@@

	% dimensions
	\newcommand{\s}{-0.5}
	\pgfmathsetmacro{\radiusats}{sqrt(1 - \s*\s)}
	\pgfmathsetmacro{\stheta}{asin(\s)}
	\pgfmathsetmacro{\minstheta}{asin(-\s)}

@@

	\definecolor{colora}{HTML}{1EBCB4}
	\definecolor{colorb}{HTML}{E57C04}
	\definecolor{colorc}{HTML}{7A5CE3}

@@

		% cut 1
		\begin{scope}[canvas is zx plane at y=\radius-\s]
			\filldraw[color=colora, thick, draw opacity=1, fill opacity=0.1] (-\margin, -\margin) -- (\margin, -\margin) -- (\margin, \margin) -- (-\margin, \margin) -- cycle;
		\end{scope}

		% cut 2
		\begin{scope}[canvas is zx plane at y=-\s]
			\filldraw[color=colorb, thick, draw opacity=1, fill opacity=0.1] (-\margin, -\margin) -- (\margin, -\margin) -- (\margin, \margin) -- (-\margin, \margin) -- cycle;
			\draw[color=colorb, thick, dashed] ({\iyz * \radiusats}, {\iyx * \radiusats}) arc[start angle={atan2(\iyx, \iyz)}, delta angle=180, radius=\radiusats cm];
			\draw[color=colorb, thick] ({-\iyz * \radiusats}, {-\iyx * \radiusats}) arc[start angle={atan2(-\iyx, -\iyz)}, delta angle=180, radius=\radiusats cm];
		\end{scope}

		% cut 3
		\begin{scope}[canvas is zx plane at y=\s]
			\filldraw[color=colorc, thick, draw opacity=1, fill opacity=0.1] (-\margin, -\margin) -- (\margin, -\margin) -- (\margin, \margin) -- (-\margin, \margin) -- cycle;
			\draw[color=colorc, thick, dashed] ({\iyz * \radiusats}, {\iyx * \radiusats}) arc[start angle={atan2(\iyx, \iyz)}, delta angle=180, radius=\radiusats cm];
			\draw[color=colorc, thick] ({-\iyz * \radiusats}, {-\iyx * \radiusats}) arc[start angle={atan2(-\iyx, -\iyz)}, delta angle=180, radius=\radiusats cm];
		\end{scope}

@@

		% empty case
		\draw[color=colora] (3.75*\xx+\xy, 3.75*\yx+\yy, 3.75*\zx+\zy) node {$s > b$};

		% northern case
		\begin{scope}[shift={(4.5*\xx-0.2*\xy, 4.5*\yx-0.2*\yy, 4.5*\zx-0.2*\zy)}, scale=0.5]
			\draw[color=colorb] (-3.5, +1.3, 0) node {$\frac{1}{2} ( a + b ) < s < b$};

@@

			% cut
			\begin{scope}[canvas is zx plane at y=-\s]
				\draw[color=colorb, thick, dashed] ({\iyz * \radiusats}, {\iyx * \radiusats}) arc[start angle={atan2(\iyx, \iyz)}, delta angle=180, radius=\radiusats cm];
				\draw[color=colorb, thick] ({-\iyz * \radiusats}, {-\iyx * \radiusats}) arc[start angle={atan2(-\iyx, -\iyz)}, delta angle=180, radius=\radiusats cm];
			\end{scope}

@@

		% southern case
		\begin{scope}[shift={(3*\xx-\xy, 3*\yx-\yy, 3*\zx-\zy)}, scale=0.5]
			\draw[color=colorc] (3.5, -0.5, 0) node {$s < \frac{1}{2} ( a + b )$};

@@

			% Bloch sphere
			\draw (\xy, \yy, \zy) arc[start angle=90, end angle={-acos(\radiusats*\iyx*\xx + \s*\yx + \radiusats*\iyz*\zx)}, radius=1 cm];
			\draw (\xy, \yy, \zy) arc[start angle=90, end angle={360-acos(-\radiusats*\iyx*\xx + \s*\yx - \radiusats*\iyz*\zx)}, radius=1 cm];
			% cut
			\begin{scope}[canvas is zx plane at y=\s]
				\draw[color=colorc, thick, dashed] ({\iyz * \radiusats}, {\iyx * \radiusats}) arc[start angle={atan2(\iyx, \iyz)}, delta angle=180, radius=\radiusats cm];
				\draw[color=colorc, thick] ({-\iyz * \radiusats}, {-\iyx * \radiusats}) arc[start angle={atan2(-\iyx, -\iyz)}, delta angle=180, radius=\radiusats cm];
			\end{scope}

>>>>

In the past few decades, there has been a growing interest in developing approaches that generalize the notion of states and physical theories in terms of probabilistic and/or entropic concepts, which stems from the underlying aim to characterize physical theories from an operational viewpoint. One basic insight is that statistical/probabilistic concepts often provide a suitable basis for this generalization, and it is in this setting where most of these approaches operate. A key problem, then, is what notion of probability should be used given that the standard measure theoretic Kolmogorovian approach does not generally work in quantum mechanics. Quasiprobabilistic approaches are also used in quantum mechanics, but it is not completely clear how exactly they relate to standard one.

@@

The starting point is that any physical law deals with experimentally reproducible relationships. This implies a physical theory must gives us the space of statistical ensembles (i.e. the collection of all possible preparation procedures) that the theory allows. As a basic requirements, ensembles must be experimentally well-defined (leading to a topological space), allow statistical mixing (leading to a convex structure) and quantify the variability of the elements of the ensembles (leading to an entropy function).\footnote{The main point of departure from GPTs is the entropy requirement. We found that a convex set is not enough to fully characterize an ensemble space, especially in the infinite dimensional case.} Additionally, though not necessary, we require having enough real valued quantities to identify each ensemble (leading to the continuous embedding into a Hausdorff locally convex topological vector space). Mathematically, under these premises, an ensembles space is a convex subset of a locally convex topological vector space. The question is, then, what measure theoretic tool can quantify the number of states (as in the classical Liouville measure) or the number of possible outcomes of a measurements (as in the dimension of orthogonal subspaces of an operator)? What measure theoretic tool can represent each ensemble? If we are to define a physically meaningful notion of probability density, both of these questions must be answered.

We will first review how measure theory is used in classical probability together with some results from Choquet theory. This will tell us that elements of a compact convex set can always be represented by a probability measure over its extreme points, the pure states. While such representation is always possible, at least in the compact case, it is not unique. Uniqueness requires the space to be a Choquet simplex, which is a generalization of the standard finite-dimensional simplex and represents the classical case. The key insight, then, an ensemble space is classical if and only if each ensembles admits a unique decomposition in terms of pure states.

We will then review some results from the use of quasi-probability measures. While the Wigner function represents the most known instance, there are different choices depending on the problem at hand. The key insight is that it is not negative probability that characterizes quasi-probability, but the lack of mutual exclusivity of the points of the space. The quasi-probability measures, in fact, are not defined over the extreme points, but over a suitable set $K$ such that the space of signed probability measures is isomorphic as a vector space to the one that embeds the ensemble space. Therefore, while signed measures are a powerful tool for calculation in specific problems, they cannot give us a general principled representation of the physical theory.

The above two approaches exhaust what can be done with additive measures, so we turn to the non-additive case. Given an ensemble $\ens$, we can ask what fraction can be produces by a mixture of a given ensembles. This leads to the fraction capacity: a sub-additive set function that, like the probability measure, is bounded to one, monotonic and continuous from above and below. Over the extreme points, the fraction capacity represents the supremum of all possible measures that can represent the same ensemble in Choquet theory. The classical case exactly corresponds to the simplex case, in which fraction capacity corresponds to a unique measure, and is therefore additive. This again solidifies the connection between classical probability, additivity of measure and mutual exclusivity.

We can also generalize a notion of count of states. While in classical statistical mechanics we start with a measure that quantifies the count of states and define the entropy using its logarithm, for the ensemble space we proceed in the opposite way: we start from the entropy and we define the count of states as the exponential of the entropy (i.e. the entropic volume). From this, we can define a state capacity, which is again a sub-additive measure. It reduces to the Liouville measure in the classical case and the dimensionality of the Hilbert space in the quantum case.

Lastly, we lay down the requirements for a non-additive generalization of probability theory. While should be in principle possible, we show that the most used non-additive integral, the Sugeno and Choquet integral, do not work for our purposes. This means that more work will be needed to develop the appropriate tools.

A final remark before we start. Too often works in the foundations of physics focus on a particular aspect without concern to a broader connection to all relevant mathematical, physical and philosophical aspects. Therefore, it is important to us that this work is a piece of a much larger puzzle, which connects cleanly and directly to  other areas, such as information geometry, functional analysis, quantum logic and so on. It is the hallmark of a true fundamental structure for physics that such connections can be easily made. Unfortunately, the nature of scientific publication requires us to publish a narrow slice of the work to fit the scope of a particular journal. Ironically, this means that the most important motivations for our approach are out of scope of this article, as we will concentrate only to the connections to measure theory.

@@

\textbf{Fraction capacity.} Let us concentrate on the following problem: given an ensemble $\ens \in \Ens$ and a (Borel) set of ensembles $A \subseteq \Ens$, what fraction of $\ens$ can be constructed by a mixture of $A$? How much of $\ens$ can be explained as coming from preparations corresponding to $A$? For example, let $\Ens$ be the space of all probability distributions for a six-sided die. Let $\ens_{123456}$ be the uniform distribution over all outcomes. Let $A_1 = \{\ens_1\}$ where $\ens_1$ represents outcome one with 100\% probability. Since we can write $\ens_{123456} = \frac{1}{6} \ens_{1} + \frac{5}{6} \ens_{23456}$ (see fig. \ref{fig:dice} for an illustration), where $\ens_{23456}$ represents a uniform distributions over the five outcomes, $1/5$ of $\ens$ can be constructed from $A$, but no more. Similarly, if $A_{12} = \{\ens_{1},\ens_{2}\}$, we can write $\ens = \frac{1}{3} \left(\frac{1}{2} \ens_1 + \frac{1}{2} \ens_2 \right)  + \frac{2}{3} \ens_{3456}$, so $1/3$ of $\ens$ can be constructed from $A$, but no more. Note that, given the uniform distribution, $1/5$ is exactly the probability for the event $A_1$ and $1/3$ the probability for event $A_{12}$. This gives the basic insight for our definitions.

\begin{figure}
	\def\xstep{0.75em}
	\begin{tikzpicture}[y=3ex, x=\xstep]
		\draw (0, 0) foreach \x in {1,...,6} { -- (\x, 0) -- (\x, -0.2) (\x, 0) } -- ++(1, 0);
		\draw plot[ybar, bar width={\xstep / 2}] coordinates{(1, 1/5) (2, 1/5) (3, 1/5) (4, 1/5) (5, 1/5) (6, 1/5)};
	\end{tikzpicture}
	$= \frac{1}{6}$
	\begin{tikzpicture}[y=3ex, x=\xstep]
		\draw (0, 0) foreach \x in {1,...,6} { -- (\x, 0) -- (\x, -0.2) (\x, 0) } -- ++(1, 0);
		\draw plot[ybar, bar width={\xstep / 2}] coordinates{(3, 1)};
	\end{tikzpicture}
	$+ \frac{5}{6}$
	\begin{tikzpicture}[y=3ex, x=\xstep]
		\draw (0, 0) foreach \x in {1,...,6} { -- (\x, 0) -- (\x, -0.2) (\x, 0) } -- ++(1, 0);
		\draw plot[ybar, bar width={\xstep / 2}] coordinates{(1, 1/5) (2, 1/4) (4, 1/5) (5, 1/5) (6, 1/5)};
	\end{tikzpicture}

	\vspace{0.3cm}
	\begin{tikzpicture}[y=3ex, x=\xstep]
		\draw (0, 0) foreach \x in {1,...,6} { -- (\x, 0) -- (\x, -0.2) (\x, 0) } -- ++(1, 0);
		\draw plot[ybar, bar width={\xstep / 2}] coordinates{(1, 1/5) (2, 1/5) (3, 1/5) (4, 1/5) (5, 1/5) (6, 1/5)};
	\end{tikzpicture}
	$= \frac{1}{3}$
	\begin{tikzpicture}[y=3ex, x=\xstep]
		\draw (0, 0) foreach \x in {1,...,6} { -- (\x, 0) -- (\x, -0.2) (\x, 0) } -- ++(1, 0);
		\draw plot[ybar, bar width={\xstep / 2}] coordinates{(3, 1/2) (4, 1/2)};
	\end{tikzpicture}
	$+ \frac{2}{3}$
	\begin{tikzpicture}[y=3ex, x=\xstep]
		\draw (0, 0) foreach \x in {1,...,6} { -- (\x, 0) -- (\x, -0.2) (\x, 0) } -- ++(1, 0);
		\draw plot[ybar, bar width={\xstep / 2}] coordinates{(1, 1/4) (2, 1/4) (5, 1/4) (6, 1/4)};
	\end{tikzpicture}
	\caption{Illustration of the mixture of distributions}\label{fig:dice}
\end{figure}

@@

	% dimensions
	\newcommand{\s}{-0.4}
	\pgfmathsetmacro{\radiusats}{sqrt(1 - \s*\s)}
	\pgfmathsetmacro{\stheta}{asin(\s)}
	\pgfmathsetmacro{\minstheta}{asin(-\s)}

@@

@@

		% cut
		\begin{scope}[canvas is zx plane at y=\s]
			\filldraw[color=blue, thick, draw opacity=1, fill opacity=0.1] (-\margin, -\margin) -- (\margin, -\margin) -- (\margin, \margin) -- (-\margin, \margin) -- cycle;
			\draw[blue, thick, dashed] ({\iyz * \radiusats}, {\iyx * \radiusats}) arc[start angle={atan2(\iyx, \iyz)}, delta angle=180, radius=\radiusats cm];
			\draw[blue, thick] ({-\iyz * \radiusats}, {-\iyx * \radiusats}) arc[start angle={atan2(-\iyx, -\iyz)}, delta angle=180, radius=\radiusats cm];
		\end{scope}

@@

		% empty case
		\draw (3.75*\xx+\xy, 3.75*\yx+\yy, 3.75*\zx+\zy) node {$s > b$};

		% northern case
		\begin{scope}[shift={(4.5*\xx-0.2*\xy, 4.5*\yx-0.2*\yy, 4.5*\zx-0.2*\zy)}, scale=0.5]
			\draw (-3.5, +1.3, 0) node {$\frac{1}{2} ( a + b ) < s < b$};

@@

			% cut
			\begin{scope}[canvas is zx plane at y=-\s]
				\draw[blue, thick, dashed] ({\iyz * \radiusats}, {\iyx * \radiusats}) arc[start angle={atan2(\iyx, \iyz)}, delta angle=180, radius=\radiusats cm];
				\draw[blue, thick] ({-\iyz * \radiusats}, {-\iyx * \radiusats}) arc[start angle={atan2(-\iyx, -\iyz)}, delta angle=180, radius=\radiusats cm];
			\end{scope}

@@

		% southern case
		\begin{scope}[shift={(3*\xx-\xy, 3*\yx-\yy, 3*\zx-\zy)}, scale=0.5]
			\draw (3.5, -0.5, 0) node {$s < \frac{1}{2} ( a + b )$};

@@

			% Bloch sphere
			\draw (\xy, \yy, \zy) arc[start angle=90, end angle={-acos(\radiusats*\iyx*\xx + \s*\yx + \radiusats*\iyz*\zx)}, radius=1 cm];
			\draw (\xy, \yy, \zy) arc[start angle=90, end angle={360-acos(-\radiusats*\iyx*\xx + \s*\yx - \radiusats*\iyz*\zx)}, radius=1 cm];
			% cut
			\begin{scope}[canvas is zx plane at y=\s]
				\draw[blue, thick, dashed] ({\iyz * \radiusats}, {\iyx * \radiusats}) arc[start angle={atan2(\iyx, \iyz)}, delta angle=180, radius=\radiusats cm];
				\draw[blue, thick] ({-\iyz * \radiusats}, {-\iyx * \radiusats}) arc[start angle={atan2(-\iyx, -\iyz)}, delta angle=180, radius=\radiusats cm];
			\end{scope}
